Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-04-02 curvature.tex

We first consider properties of the uniform (a.k.a.\ isotropic) random walk on the rSPR graph.
Assume that the random walk begins at a tree $T$.
The uniform random walk selects a tree uniformly at random from $N(T)$.
\cuttable{We denote the curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.

@@

We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.

@@

Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees'', and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}

>>>>

We first consider properties of the uniform (a.k.a.\ isotropic) random walk on the rSPR graph.
Assume that the random walk begins at a tree $T$.
The uniform random walk selects a tree uniformly at random from $N(T)$.
\cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.

@@

We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.

@@

Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
